\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-27 paper.tex
Tweak

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% TODO more?
mathematically defined~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
In this Griffiths encoding, as well as nodes marking sampling events,
we have two additional node types in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

>>>>

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where such concrete details of an ARG data structure
(the encoding) are
% WIP - it's surprisingly hard to find papers where an ARG is defined
% concretely
mathematically defined~\citep[e.g.][]{
gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
In this Griffiths encoding, as well as nodes marking sampling events,
we have two additional node types in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints. These breakpoints
are used to construct the local trees.
This is done by tracing rootwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
